\chapter{Processes}
\label{processes}

A \textbf{process} corresponds to one \textit{thread of
  control}. Erlang permits very large numbers of concurrent processes
each executing like it had an own virtual processor. When a process
executing a \texttt{function A} calls another \texttt{function B}, it
will wait until \texttt{function B} is finished and then retrieve its
result. If instead it \textit{spawns} another process executing
\texttt{function B}, both will continue in parallel
(concurrently). \texttt{function A} will not wait for \texttt{function
  B} and the only way they can communicate is through \textit{message
  passing}.

Erlang processes are light-weight with small memory footprint, fast to
create and terminate and the scheduling overhead is low. A
\textbf{process identifier}, \texttt{Pid}, identifies a process. The
BIF \texttt{self/0} returns the \texttt{Pid} of the process itself.


\section{Process creation}
A process is created using the BIF \texttt{spawn/3}.

\begin{erlang}
spawn(Module, Func, [Expr1, ..., ExprN])
\end{erlang}

\texttt{Module} should evaluate to a module name and \texttt{Func} to
a function name in that module. The list of \texttt{Exprs} are the
arguments to the function. \texttt{spawn} creates a new process and
returns the process identifier, \texttt{Pid}. The new process starts
by executing

\begin{erlang}
Module:Func(Expr1, ..., ExprN)
\end{erlang}

The function \texttt{Func} has to be exported even if it is spawned by
another function in the same module. There are other spawn BIFs, for
example \texttt{spawn/4} for spawning a process at another node.


\section{Registered processes}
A process can be associated with a name. The name must be an atom and
is automatically unregistered if the process terminates. Only static
(cyclic) processes should be registered.

\begin{center}
\begin{tabular}{|>{\raggedright}p{117pt}|>{\raggedright}p{204pt}|}
\hline
\multicolumn{2}{|p{321pt}|}{Name registration BIFs}\tabularnewline
\hline
\texttt{register(Name, Pid)}  & Associates the atom Name with the process Pid\tabularnewline
\hline
\texttt{registered()}  & Returns a list of names which have been registered \tabularnewline
\hline
\texttt{whereis(Name)}  & Returns the Pid registered under Name or undefined if the name
is not registered\tabularnewline
\hline
\end{tabular}
\end{center}


\section{Process communication}
Processes communicate by sending and receiving
\textbf{messages}. Messages are sent using the send operator
(\texttt{!}) and are received using \texttt{receive}. Message passing
is asynchronous and reliable, i.e. the message is guaranteed to
eventually reach the recipient, provided that the recipient exists.


\subsection{Send}

\begin{erlang}
Pid ! Expr
\end{erlang}

The send (\texttt{!}) operator sends the value of \texttt{Expr} as a
message to the process specified by \texttt{Pid} where it will be
placed last in its \textbf{message queue}. The value of \texttt{Expr}
is also the return value of the (\texttt{!}) expression. \texttt{Pid}
must evaluate to a process identifier, a registered name or a tuple
\texttt{\{Name,Node\}}, where \texttt{Name} is a registered process at
\texttt{Node} (see chapter \ref{distribution}). The message sending operator
(\texttt{!}) never fails, even if it addresses a non-existing process.


\subsection{Receive}

\begin{erlang}
receive
    Pattern1 [when GuardSeq1] ->
        Body1;
    ...
    PatternN [when GuardSeqN] ->
        BodyN                   % Note no semicolon (;) before end
end
\end{erlang}

This expression receives messages sent to the process using the send
operator (\texttt{!}). The patterns \texttt{PatternX} are sequentially
matched against the first message in time order in the message queue,
then the second, etc. If a match succeeds and the optional guard
sequence \texttt{GuardSeqX} is true, then the message is removed from
the message queue and the corresponding \texttt{BodyX} is
evaluated. It is the order of the pattern clauses that decides the
order in which messages will be received prior to the order in which
they have arrived. This is called \textit{selective receive}. The
return value of \texttt{BodyX} is the return value of the receive
expression.

\texttt{receive} never fails. The process is suspended, possibly
indefinitely, until a message arrives that matches one of the patterns
and with a true guard sequence.


\begin{erlang}
wait_for_onhook() ->
    receive
        onhook ->
            disconnect(),
            idle();
        {connect, B} ->
            B ! {busy, self()},
            wait_for_onhook()
    end.
\end{erlang}


\subsection{Receive with timeout}

\begin{erlang}
receive
    Pattern1 [when GuardSeq1] ->
        Body1;
        ...;
    PatternN [when GuardSeqN] ->
        BodyN
after
    ExprT ->
        BodyT
end
\end{erlang}

\texttt{ExprT} should evaluate to an integer between \texttt{0} and
\texttt{16\#ffffffff} (the value must fit in 32 bits). If no matching
message has arrived within \texttt{ExprT} milliseconds, then
\texttt{BodyT} will be evaluated and its return value becomes the
return value of the receive expression.

\begin{erlang}
wait_for_onhook() ->
    receive
        onhook ->
            disconnect(),
            idle();
        {connect, B} ->
            B ! {busy, self()},
            wait_for_onhook()
    after
        60000 ->
            disconnect(),
            error()
    end.
\end{erlang}

A \texttt{receive...after} expression with no branches can be used to
implement simple timers.

\begin{erlang}
receive
after
    ExprT ->
        BodyT
end
\end{erlang}

\begin{center}
\begin{tabular}{|>{\raggedright}p{47pt}|>{\raggedright}p{273pt}|}
\hline
\multicolumn{2}{|p{321pt}|}{Two special cases for the timeout value \texttt{ExprT}}\tabularnewline
\hline
\texttt{infinity} & This is equivalent to not using a timeout and can be useful for timeout
values that are calculated at run-time\tabularnewline
\hline
\texttt{0} & If there is no matching message in the mailbox, the timeout will occur immediately\tabularnewline
\hline
\end{tabular}
\end{center}


\section{Process termination}
\label{processes:termination}
A process always terminates with an \textbf{exit reason }which may be
any term. If a process terminates normally, for instance if it has run
to the end of its code, then the reason is the atom \texttt{normal}. A process
can terminate itself by calling one of the following BIFs.

\begin{erlang}
exit(Reason)

erlang:error(Reason)

erlang:error(Reason, Args)
\end{erlang}

A process terminates with exit reason \texttt{\{Reason,Stack\}} when a
run-time error occurs.

A process may also be terminated if it receives an exit signal with
another exit reason than \texttt{normal} (see section
\ref{processes:recvexitsignals}).


\section{Process links}
\label{processes:links}
Two processes can be \textbf{linked} to each other. Links are
bidirectional and there can only be one link between two processes. A
process with \texttt{Pid1} can link to a process with \texttt{Pid2}
using the BIF \texttt{link(Pid2)}. The BIF \texttt{spawn\_link(Module,
  Func, Args)} spawns and links a process in one atomic operation.

A link can be removed using the BIF \texttt{unlink(Pid)}.


\subsection{Error handling between processes}
When a process terminates it will send \textbf{exit signals} to all
processes that it is linked to. These in turn will also be terminated
\textit{or handle the exit signal in some way}. This feature can be
used to build hierarchical program structures where some processes are
supervising other processes, for example restarting them if they
terminate abnormally.


\subsection{Sending exit signals}
\label{processes:sendexitsignals}
A process always terminates with an exit reason which is sent as an
exit signal to all linked processes. The BIF \texttt{exit(Pid,
  Reason)} sends an exit signal with the exit reason Reason to
\texttt{Pid}, without affecting the calling process.


\subsection{Receiving exit signals}
\label{processes:recvexitsignals}
If a process receives an exit signal with an exit reason other than
normal it will also be terminated and will send exit signals with the
same exit reason to its linked processes. An exit signal with reason
normal is ignored. This behaviour can be changed using the BIF
\texttt{process\_flag(trap\_exit, true)}.

Then the process is able to \textbf{trap exits}. This means that an
exit signal will be transformed into a message \texttt{\{'EXIT',
  FromPid, Reason\}} which is put into the message queue and can be
handled by the process like a regular message using receive.

However, a call to the BIF \texttt{exit(Pid, kill)} unconditionally
terminates the process \texttt{Pid} regardless whether it is able to
trap exit signals or not.


\section{Monitors}
A process \texttt{Pid1} can create a \textbf{monitor} for
\texttt{Pid2} using the BIF

\begin{erlang}
erlang:monitor(process, Pid2)
\end{erlang}

which returns a reference \texttt{Ref}. If \texttt{Pid2} terminates
with exit reason \texttt{Reason}, a message as follows will be sent to
\texttt{Pid1}.

\begin{erlang}
{'DOWN', Ref, process, Pid2, Reason}
\end{erlang}

If \texttt{Pid2} does not exist, the \texttt{'DOWN'} message is sent
immediately with \texttt{Reason} set to \texttt{noproc}. Monitors are
unidirectional in that if \texttt{Pid1} monitors \texttt{Pid2} then it
will receive a message when \texttt{Pid2} dies but \texttt{Pid2} will
\textbf{not} receive a message when \texttt{Pid1} dies. Repeated calls
to \texttt{erlang:monitor(process, Pid)} will create several,
independent monitors and each one will send a \texttt{'DOWN'} message
when \texttt{Pid} terminates.

A monitor can be removed by calling \texttt{erlang:demonitor(Ref)}. It
is possible to create monitors for processes with registered names,
also at other nodes.


\section{Process priorities}
The BIF \texttt{process\_flag(priority, Prio)} defines the priority of
the current process. \texttt{Prio} may have the value \texttt{normal},
which is the default, or \texttt{high} or \texttt{low} or \texttt{max}. This should be
used very sparingly.


\section{Process dictionary}
\label{processes:dicts}
Each process has its own process dictionary which is a list of
\texttt{\{Key, Value\}} terms.

\begin{center}
\begin{tabular}{|>{\raggedright}p{79pt}|>{\raggedright}p{247pt}|}
\hline
\multicolumn{2}{|p{326pt}|}{Process dictionary BIFs}\tabularnewline
\hline
\texttt{put(Key, Value)} & Saves the Value under the Key or replaces an older value\tabularnewline
\hline
\texttt{get(Key)} & Retrieves the Value stored under Key or undefined\tabularnewline
\hline
\texttt{get()} & Returns the entire process dictionary as a list of \{Key, Value\} terms\tabularnewline
\hline
\texttt{get\_keys(Value)} & Returns a list of keys that have the value Value\tabularnewline
\hline
\texttt{erase(Key)} & Deletes \{Key, Value\}, if any, and returns Key\tabularnewline
\hline
\texttt{erase()} & Returns the entire process dictionary and deletes it\tabularnewline
\hline
\end{tabular}
\end{center}

Process dictionaries could be used to keep global variables within an application but the extensive use of them is usually regarded as poor programming style.
