\chapter{Distributed Erlang}
\label{distribution}

A \textbf{distributed Erlang system} consists of a number of Erlang
runtime systems communicating with each other. Each such runtime
system is called a \textbf{node}.  Nodes can reside on the same
host or on different hosts connected through a network.  The
standard distribution mechanism is implemented using TCP/IP sockets
but other mechanisms can also be implemented.

Message passing between processes on different nodes, as well as links
and monitors, is transparent when using \texttt{Pid}s.  However, registered
names are local to each node.  A registered process at a particular
node is referred to as \texttt{\{Name,Node\}}.

The Erlang Port Mapper Daemon \textbf{epmd} is automatically started
on every host where an Erlang node is started.  It is responsible for
mapping the symbolic node names to machine addresses.


\section{Nodes}
A \textbf{node} is an executing Erlang runtime system which has been
given a name, using the command line flag \texttt{-name} (long name)
or \texttt{-sname} (short name).

The format of the node name is an atom \texttt{Name@Host} where
\texttt{Name} is the name given by the user and \texttt{Host} is the
full host name if long names are used, or the first part of the host
name if short names are used. \texttt{node()} returns the name of the
node. Nodes using long names cannot communicate with nodes using
short names.


\section{Node connections}
The nodes in a distributed Erlang system are fully connected. The
first time the name of another node is used, a connection attempt to
that node will be made. If a node A connects to node B, and node B has
a connection to node C, then node A will also try to connect to node
C. This feature can be turned off using the command line flag:

\texttt{~~~~-connect\_all false}

If a node goes down, all connections to that node are removed. The
BIF:

\begin{erlang}
erlang:disconnect(Node)
\end{erlang}

disconnects \texttt{Node}. The BIF \texttt{nodes()} returns the list of
currently connected (visible) nodes.


\section{Hidden nodes}
It is sometimes useful to connect to a node without also connecting to
all other nodes. For this purpose, a \textbf{hidden node} may be
used. A hidden node is a node started with the command line flag
\texttt{-hidden}.  Connections between hidden nodes and other nodes must be set
up explicitly. Hidden nodes do not show up in the list of nodes
returned by \texttt{nodes()}. Instead, \texttt{nodes(hidden)} or \texttt{nodes(connected)} must
be used. A hidden node will not be included in the set of nodes that
the module global keeps track of.

A \textbf{C node} is a C program written to act as a hidden node in a
distributed Erlang system. The library \texttt{erl\_interface}
contains functions for this purpose.


\section{Cookies}
Each node has its own \textbf{magic cookie}, which is an atom. The
Erlang network authentication server (auth) reads the cookie in the
file \texttt{\$HOME/.erlang.cookie}. If the file does not exist, it
will be created with a random string as content.

% FRMB CHECK: the implication here is that "erlang:set_cookie(node(), Cookie)"
% sets *this* node's cookie, as well as the cookie that will be used to connect
% to other nodes, if not explicitly set otherwise.

The permissions of the file must be set to octal 400 (read-only
by user).  The cookie of the local node may also be set using the BIF
\texttt{erlang:set\_cookie(node(), Cookie)}.

The current node is only allowed to communicate with another node
\texttt{Node2} if it knows its cookie.  If this is different from the current node (whose cookie will
be used by default) it must be explicitly set with the BIF \texttt{erlang:set\_cookie(Node2, Cookie2)}.


\section{Distribution BIFs}

\begin{center}
\begin{tabular}{|>{\raggedright}p{156pt}|>{\raggedright}p{170pt}|}
\hline
\multicolumn{2}{|p{326pt}|}{Distribution BIFs}\tabularnewline
\hline
\texttt{node()}  & Returns the name of the current node. Allowed in guards\tabularnewline
\hline
\texttt{is\_alive()}  & Returns true if the runtime system is a node and can connect to
other nodes, false otherwise\tabularnewline
\hline
\texttt{erlang:get\_cookie()}  & Returns the magic cookie of the current node\tabularnewline
\hline
\texttt{set\_cookie(Node, Cookie)} & Sets the magic cookie used when connecting to \texttt{Node}.
If \texttt{Node} is the current node, \texttt{Cookie} will be used when connecting to all new nodes\tabularnewline
\hline
\texttt{nodes()}  & Returns a list of all visible nodes to which the current node is connected
to\tabularnewline
\hline
\texttt{nodes(connected\textbar{}hidden)}  & Returns a list not only of visible nodes,
but also hidden nodes and previously known nodes, etc. \tabularnewline
\hline
\texttt{monitor\_node(Node,}\break\texttt{\phantom{xxxx}true\textbar{}false)}  & Monitors the status of \texttt{Node}. A message
\texttt{\{nodedown, Node\}} is received if the connection to it is lost\tabularnewline
\hline
\texttt{node(Pid\textbar{}Ref\textbar{}Port)}  & Returns the node where the argument is
located\tabularnewline
\hline
\texttt{erlang:disconnect\_node(Node)}  & Forces the disconnection of \texttt{Node}\tabularnewline
\hline
\texttt{spawn[\_link\textbar{}\_opt](Node,}\break\texttt{\phantom{xxxx}Module, Function, Args)}  & Creates a process
at a remote node\tabularnewline
\hline
\texttt{spawn[\_link\textbar{}\_opt](Node, Fun)}  & Creates a process at a remote node\tabularnewline
\hline
\end{tabular}
\end{center}


\section{Distribution command line flags}

\begin{center}
\begin{tabular}{|>{\raggedright}p{102pt}|>{\raggedright}p{224pt}|}
\hline
\multicolumn{2}{|p{326pt}|}{Distribution command line flags}\tabularnewline
\hline
\texttt{-connect\_all false}  & Only explicit connection set-ups will be used\tabularnewline
\hline
\texttt{-hidden}  & Makes a node into a hidden node\tabularnewline
\hline
\texttt{-name Name}  & Makes a runtime system into a node, using long node names\tabularnewline
\hline
\texttt{-setcookie Cookie}  & Same as calling \linebreak{}
\texttt{erlang:set\_cookie(node(), Cookie))}\tabularnewline
\hline
\texttt{-sname Name}  & Makes a runtime system into a node, using short node names\tabularnewline
\hline
\end{tabular}
\end{center}


\section{Distribution modules}
There are several modules available which are useful for distributed programming:

\begin{center}
\begin{tabular}{|>{\raggedright}p{93pt}|>{\raggedright}p{233pt}|}
\hline
\multicolumn{2}{|p{326pt}|}{Kernel modules useful for distribution}\tabularnewline
\hline
\texttt{global}  & A global name registration facility\tabularnewline
\hline
\texttt{global\_group}  & Grouping nodes to global name registration groups\tabularnewline
\hline
\texttt{net\_adm}  & Various net administration routines\tabularnewline
\hline
\texttt{net\_kernel}  & Erlang networking kernel\tabularnewline
\hline
\multicolumn{2}{|p{326pt}|}{STDLIB modules useful for distribution}\tabularnewline
\hline
\texttt{slave}  & Start and control of slave nodes\tabularnewline
\hline
\end{tabular}
\end{center}
